### 2018-04-18 Writing/disparity_review.tex
updated the metrics section

[INCLUDE WILLS FIGURE HERE]

@@

After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of it to test hypotheses.
Because of these reduction in dimensionality, these metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
With that in mind, the choice of the metric should be essential to summarise the aspect of the morphospace that is the most related to the question.
In an unidimensional distributions (e.g. histograms) different metrics will readily reflect different aspect of the data.
For example, metrics can be used to reflect the ``size'' of the distribution (e.g. the range, quantiles or variance) and some other will reflect the most likely value (the central tendency - e.g. the mean, median or mode).
Among these metrics, some will have more preferable properties than other such as sensitivity to outliers (range, mean or mode \textit{vs.} quantiles, variance or median) and will thus be more or less appropriate to the different questions.

In multidimensionality, this high availability of unidimensional descriptors comes with equally numerous advantages and caveats.
If the question refers to ``size'' of a group in the morphospace (e.g. ``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriated (e.g. volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g. ``does group A occupy the same space as group B?''), metrics relating to the distance of the elements to a point are most appropriated (e.g. distance from a fixed point - e.g. the opposite group).
Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g. ``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriated (e.g. nearest neighbour distance, pairwise distances, etc.)

When looking at this positively, any metric can be suitable to any question, when looking at it negatively, it is important to keep in mind several caveats that can apply in any case:
First, it is important to relate the metrics to the space property.
For example, measuring the sum of variances of the space's dimensions does not require the co-variance element between the axis in a PCA but only when the full space is considered: if only a subset is measured, co-variance should be added to the sum of variances.
Second, some metrics automatically relate to the properties of the morphospace.
For example, range or hull based metrics are really sensitive to outliers and thus biased based on the sample size.

Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.

Best practices:
\begin{itemize}
\item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper. The definition can be in plain English: e.g. ``we measure disparity as the sum of the variance of each axis of the ordinated matrix''; or mathematically: e.g. ``we measured disparity as $\sum\sigma_i$ where $\sigma_i$ is the variance of the each ordinated axis $i$.''.
\item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true). We therefore suggest to always plot some dimensions of the space (see Visualisation section \ref{visualisation}) and try multiple metrics.
\end{itemize}

>>>>

[INCLUDE WILLS FIGURE SOMWHERE AROUND]

@@

After building and visualising the morphospace, one can use disparity metrics (or indices) as a one dimensional summary of the space and to test hypotheses.
As any summary metric and because of this reduction in dimensionality, disparity metrics are bound to only reflect some aspect of the morphospace and can never encompass its whole complexity.
It is often beneficial to use more than one such a metric to summarise the different aspects of interest of the morphospace (e.g. it's width, it's spread, etc.).

The choice of the disparity metric is thus essential to the disparity analysis and should always be driven by the question.
When considering only one dimension, disparity metrics can be used to reflect the ``size'' of the distribution (e.g. the range, quantiles or variance) and some other will reflect the most likely value (the central tendency - e.g. the mean, median or mode).
Among these metrics, some will have more preferable properties than other such as sensitivity to outliers (range, mean or mode \textit{vs.} quantiles, variance or median) and will thus be more or less appropriate to the different questions and to the actual properties distribution.
For example, if the data collected is likely to contain outliers, but is fairly normally distributed, one should prefer the median to the mean.

With multidimensionality, this high availability of unidimensional descriptors comes with an equally numerous advantages and caveats and again, the question can help determining which metric would be the most appropriate.
For example, if the question refers to ``size'' of a group in the morphospace (e.g. ``does group A occupy as much space as group B?''), metrics relating to the spread of the group in the morphospace are the most appropriated (e.g. volume [CITE], variance [CITE], range [CITE], distance from the centroid [CITE]).
Conversely, if the question refers to the ``position'' of a group in the morphospace (e.g. ``does group A occupy the same space as group B?''), metrics relating to the distance between the elements within a group and a fixed point in the morphospace are most appropriated.
Finally, one might also be interested in the density of elements in some regions of the morphospace (e.g. ``is group A more dense than group B?''), metrics relating to the pairwise distances between elements are most appropriated (e.g. nearest neighbour distance, pairwise distances, etc.).

Additionally to the desired properties of what the metrics should capture, it is important to also keep in mind the mathematical properties of the metrics and their associated caveats.

\begin{itemize} %TG: not sure how to introduce this list. Do we even need a list? I can shorten these examples and discuss them more in my paper-project on disparity metrics?
    \item For example, measuring the full sum of the variance of each dimensions of the space does not require to add the co-variance between the axis in a ordinated space using a PCA.
    This is not true however in other mathematical spaces or (even in a PCA) when not all dimensions or all the elements are considered (e.g. when only looking at a group within the morphospace, the sum of variances is not mathematically equal to the sum of the variance of each dimensions in the space).

    \item Furthermore, multidimensional space have some counter-intuitive properties that needs to be taken into account such as the curse of dimensionality [CITE].
    For example, product based metrics (e.g. product of ranges, hypervolume, hypercube, etc.) will tend towards 0 fairly quickly for spaces with more than 6 to 10 dimensions.

    \item Some other types of metrics are also extremely sensitive to outliers and can be easily biased by sample size such as range or convex hull based metrics.

\end{itemize}

%TG: again, I feel this section is already fairly long, I might include this bit in another paper
% Of course some other aspects could be of interest (or more than one) and thus some other metrics would be more appropriated.
% For example, differential changes (in direction or relative magnitude) of sum of ranges and sum of variances can be indicative of different modes of extinction selectivity (e.g., Korn et al 2013, Evolution).
% Furthermore, one future avenue that could be explored is to go beyond single valued disparity metrics and rather express them in terms of distributions.
% For example, when the metric is expressed as the median distance from centroids [e.g. in CITE], it could be informative as well to express it as a full distribution.
% This would allow to both visualise the variation at a glance and use this variation in statistical tests for comparing these distributions.

% Best practices:
% \begin{itemize}
% \item Facing the amount of different ways to measure disparity and regardless which metric is the most suited to the question, it is crucial to clearly and mathematically define the metric in a disparity paper. The definition can be in plain English: e.g. ``we measure disparity as the sum of the variance of each axis of the ordinated matrix''; or mathematically: e.g. ``we measured disparity as $\sum\sigma_i$ where $\sigma_i$ is the variance of the each ordinated axis $i$.''.
% \item It is important to keep in mind that different metrics can lead to the same results even though the described spaces are different (the opposite is \textit{not} true). We therefore suggest to always plot some dimensions of the space (see Visualisation section \ref{visualisation}) and try multiple metrics.
% \end{itemize}
